To be FAIR: Theory Specification Needs an Update

### 2025-06-06 manuscript.Rmd
update 2025-06-06 10:16:10.132865

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "7"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

```{r tabmeehl}
data.frame(
  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
  "X-Interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
) |>
papaja::apa_table()
```

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
  - id            : "7"
    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers^[[As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.]{#faircap}].
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
[FAIR theory has the potential to improve the efficiency of scholarly communication and]{#test2}
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

@@

```{r tabmeehl}
tab_meehl <- data.frame(
  "Property" = c("1) Ontology", "2) Causal connections", "3-8) Functional Form", "9) Numerical Value"),
  "X-Interoperability" = c("Variable selection", "Model specification, covariate selection, causal inference", "Deriving specific hypotheses", "Simulating data"), check.names = FALSE
)
write.csv(tab_meehl, "tab_meehl.csv")
papaja::apa_table(tab_meehl)
```

### 2025-06-06 manuscript.tex
update 2025-06-06 10:16:10.132865

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9211}
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 7b710ba5. The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, dept. Methodology \& Statistics\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Max Planck School of Cognition, Leipzig, Germany\\\textsuperscript{5} University of Amsterdam, the Netherlands\\\textsuperscript{6} Ludwig-Maximilians-Universität München, Germany\\\textsuperscript{7} Department of Psychology, MSB Medical School Berlin, Berlin, Germany}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016) for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta science, theory formation, open science\newline\indent Word count: 9245}
\usepackage{lineno}

@@

This is a preprint paper, generated from Git Commit \# 7d4ab9cc. The first and senior author positions were fixed, and remaining author positions randomized using \texttt{set.seed(6999982);\ sample(c("Aaron",\ "Felix",\ "Noah",\ "Maximilian"))}.

@@

\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University department of Methodology \& Statistics, Tilburg, The Netherlands\\\textsuperscript{2} Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany\\\textsuperscript{3} Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany\\\textsuperscript{4} Max Planck School of Cognition, Leipzig, Germany\\\textsuperscript{5} University of Amsterdam, Amsterdam, The Netherlands\\\textsuperscript{6} Ludwig-Maximilians-Universität München, München, Germany\\\textsuperscript{7} MSB Medical School Berlin Department of Psychology, Berlin, Germany}

@@

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable (M. D. Wilkinson et al., 2016) for both humans and computers\footnote{\phantomsection\label{testone}{As the colloquial use of these terms differs in important ways from their definition according to the FAIR principles, we capitalize these terms when we refer to a specific FAIR principle.}}.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artifacts more open, such as research software (Lamprecht et al., 2019) and computational workflows (S. R. Wilkinson et al., 2024).
This paper argues that the FAIR principles can similarly advance effective and transparent scholarly communication about theory.
To this end, we introduce ``FAIR theory'':
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
\phantomsection\label{test2}{FAIR theory has the potential to improve the efficiency of scholarly communication and}
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences and beyond.
